---
title: Analyse results of SourceTracker-based microbial matching
author: David Wilkins
date: Last updated `r lubridate::today()`
output:
  html_document:
    toc: true
    theme: readable
---

```{r global_options, include = FALSE}
library(knitr)
opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r message = FALSE}
library(tidyverse)
library(printr)
library(fs)
library(ggfittext)
library(vegan)
library(wilkoxmisc)
library(ggtext)
set.seed(2)
```

The purpose of this analysis is to analyse the results of the
SourceTracker-based microbial matching. This is part of my broader aims of
comparing the accuracy of different matching methods and of testing for an
effect of sampling delay on matching accuracy over the short time period of
this study.

## Load and tidy the SourceTracker source predictions

```{r}
metadata <- read_tsv("../metadata/metadata.tsv") %>%
  rename_all(~ str_c("sink_", .))

source_predictions <- tibble(sp_dir = dir_ls("source_predictions")) %>%
  mutate(source_timepoint = path_file(sp_dir)) %>%
  mutate(sp_path = path(sp_dir, "sink_predictions", ext = "txt")) %>%
  mutate(sp = map(sp_path, read_tsv)) %>%
  select(source_timepoint, sp) %>%
  unnest(sp) %>%
  rename(sink_sample = SampleID) %>%
  gather("source_location", "proportion", -source_timepoint, -sink_sample) %>%
  left_join(metadata) %>%
  type_convert()
```

## Visualise source predictions for all sink sites and time point

I'll plot an overview of the source predictions for each sink site and time
point, with the sources grouped by same location, different location and
unknown.

```{r}
source_predictions %>%
  mutate(source = case_when(
    source_location == "Unknown" ~ "Unknown",
    source_location == sink_location ~ "Same location",
    ! source_location == sink_location ~ "Different location",
    TRUE ~ NA_character_)) %>%
  count(source_timepoint, sink_sample, sink_site, source, wt = proportion, 
        name = "proportion") %>%
  mutate_at("sink_site", str_to_sentence) %>%
  mutate_at("source", fct_relevel, "Same location", "Different location", "Unknown") %>%
  ggplot(aes(x = sink_sample, y = proportion, fill = source)) +
    geom_col(position = "stack") +
    facet_grid(source_timepoint ~ sink_site, scales = "free_x") +
    labs(title = "SourceTracker predicted sources for all surface sites",
         subtitle = "Faceted by source time point",
         x = "Surface sample",
         y = "Predicted source proportion",
         fill = "Source") +
    scale_fill_manual(values = c("#8dd3c7", "#fb8072", "grey")) +
    scale_y_continuous(breaks = c(0, 0.5, 1)) +
    theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

Impression:

- For the household surface sites, same-location skin was the majority source for
  most time points.
- This does not hold for public surface sites, where different-location or
  unknown sources predominated (though this does not necessarily mean that the
  same location was not the single major predicted source).
- There is no clearly visible effect of time point.
- Within the bed headboard site, and to a lesser extend the door knob site,
  there is clearly some heterogeneity within the sink samples; this likely
  relates to the location from which these samples were taken.

## Matching accuracy

I'll now proceed to determining the matching accuracy with the SourceTracker
method. I'll use my usual criterion for determining a match: the single
non-Unknown source with the highest predicted source proportion will be awarded
the match for a given sink sample. If the matching source is from the same
location as the sink sample, the match is considered accurate.

I want to answer the following questions regarding the matching accuracy:

- What is the overall accuracy rate?
- What is the accuracy rate broken down by sink site?
- How does sampling delay (time elapsed between taking a reference sample and
  taking a query sample; a negative value means the query sample was taken
  before the reference sample) affect matching accuracy, both globally and by
  site?
- How does the time of day at which the query and reference was taken affect
  matching accuracy (diurnality effect)?

### Determining match and accuracy for each query

First, I'll count the number of matching attempts with ties.

```{r}
source_predictions %>%
  filter(! source_location == "Unknown") %>%
  group_by(source_timepoint, sink_sample) %>%
  mutate(winner = proportion == max(proportion)) %>%
  ungroup() %>%
  filter(winner) %>%
  count(source_timepoint, sink_sample) %>%
  filter(n > 1) %>%
  select(sample = sink_sample) %>%
  left_join(read_tsv("../metadata/metadata.tsv")) %>%
  mutate_at(c("site", "location"), str_to_sentence) %>%
  ggplot(aes(x = site, fill = location)) +
    geom_bar(position = "stack") +
    labs(title = "SourceTracker matching attempts with ties, by sink site and location",
         x = "Site",
         y = "Attempts",
         fill = "Location")
```

There were 133 ties. All were for park/campus handrail sinks. I want to
visualise the proportion of all runs this represents, to better understand the
effect of removing these runs.

```{r}
source_predictions %>%
  filter(! source_location == "Unknown") %>%
  group_by(source_timepoint, sink_sample) %>%
  mutate(winner = proportion == max(proportion)) %>%
  ungroup() %>%
  filter(winner) %>%
  count(source_timepoint, sink_sample) %>%
  mutate(tie = n > 1) %>%
  select(sample = sink_sample, source_timepoint, tie) %>%
  left_join(read_tsv("../metadata/metadata.tsv")) %>%
  select(source_timepoint, site, location, tie) %>%
  mutate_at(c("site", "location"), str_to_sentence) %>%
  mutate(location = str_remove(location, "Residence ")) %>%
  ggplot(aes(x = location, fill = tie)) +
    geom_bar(position = "stack") +
    facet_grid(source_timepoint ~ site) +
    labs(title = "SourceTracker matching attempts with ties",
         subtitle = "By source timepoint and sink site and location",
         x = "Residence",
         y = "Attempts",
         fill = "Tied?") +
    scale_y_continuous(breaks = c(0, 10, 20))
```

Fortunately, they are a small proportion overall and spread over three
different residences. Still, I'd like to better understand why this is
happening, especially as the original dataset did not have so many ties.

I'll make a list of the runs that produced ties, and look more closely at the
source predictions for these runs.

```{r}
tied_attempts <- source_predictions %>%
  filter(! source_location == "Unknown") %>%
  group_by(source_timepoint, sink_sample) %>%
  mutate(winner = proportion == max(proportion)) %>%
  ungroup() %>%
  filter(winner) %>%
  count(source_timepoint, sink_sample) %>%
  filter(n > 1) %>%
  select(-n)

tied_attempts %>%
  left_join(source_predictions) %>%
  distinct(source_location, proportion)
```

Essentially, for these runs there was zero predicted contribution from any
known source. I smell a rat here.

```{r}
tied_attempts %>%
  count(sink_sample)
```

It doesn't seem to be an issue with the sink samples themselves, as they didn't
all tie all runs but only some runs (i.e. some source timepoints).

```{r}
tied_attempts %>%
  distinct(sink_sample)
```

There are only 13 sink samples in total that failed. From eyeballing the
species abundances for these sample, I didn't notice anything out of the
ordinary - they seemed to have plenty of species shared with the skin (source)
samples for the timepoints that failed.

Just for the sake of it, I'm going to ordinate the Bray-Curtis dissimilarities
for all the park/campus handrail samples and see if these ones cluster away.

```{r}
abundances <- read_tsv("../abundances/species_abundances_decontaminated.tsv")

abundances %>%
  filter(site == "park/campus handrail") %>%
  select(sample, species, abundance) %>%
  arrange(sample) %>%
  spread(species, abundance, fill = 0) %>%
  column_to_rownames("sample") %>%
  as.matrix %>%
  vegdist() %>%
  metaMDS(try = 100) %>%
  .$points %>% 
  as.data.frame %>% 
  rownames_to_column("sample") %>% 
  as_tibble() %>%
  mutate(tied = sample %in% tied_attempts$sink_sample) %>%
  ggplot(aes(x = MDS1, y = MDS2, colour = tied)) +
    geom_point() +
    labs(title = "nMDS of all park/campus handrail samples",
         subtitle = "Based on Bray-Curtis dissimilarities",
         colour = "Tied?")
```

Yes, these samples are all peripheral to the central cluster. I'm going to
compare their taxonomic profiles.

```{r}
abundances %>%
  filter(site == "park/campus handrail") %>%
  mutate(tied = sample %in% tied_attempts$sink_sample) %>%
  mutate(species = fct_lump(species, n = 11, w = abundance)) %>%
  mutate_at("species", fct_relevel, "Other", after = Inf) %>%
  count(tied, location, timepoint, species, wt = abundance, name = "abundance") %>%
  mutate(tied = if_else(tied, "Tied", "Not tied")) %>%
  mutate_at("species", binom) %>%
  mutate_at("location", str_to_sentence) %>%
  ggplot(aes(x = timepoint, y = abundance, fill = species)) +
    geom_col(position = "stack") +
    facet_grid(tied ~ location, scales = "free_x") +
    scale_fill_brewer(palette = "Set3") +
    labs(title = "Taxonomic composition of park/campus handrail sites",
         x = "Time point",
         y = "Relative abundance (%)",
         fill = "Species") +
    theme(legend.text = element_markdown())
```

With the exception of the Residence 4 sample (which was dominated by a single
*Halomonas* species – a likely contaminant), there's nothing strikingly
different about the tied samples.

```{r}
matches <- source_predictions %>%
  filter(! source_location == "Unknown") %>%
  group_by(source_timepoint, sink_sample) %>%
  mutate(winner = proportion == max(proportion)) %>%
  ungroup() %>%
  filter(winner) %>%
  add_count(source_timepoint, sink_sample) %>%
  filter(n == 1) %>%
  select(-n) %>%
  mutate(accurate = source_location == sink_location) %>%
  select(-proportion)
```

I'll write this to file for later use.

```{r}
write_tsv(matches, "st_matches.tsv")
```

## Overall and site accuracy rates

The overall accuracy rate:

```{r}
matches %>%
  summarise(accuracy = sum(accurate) / n())
```

This is even a little higher than I would expect. Now I'll calculate the rate
broken down by type:

```{r}
matches %>%
  group_by(sink_type) %>%
  summarise(accuracy = sum(accurate) / n())
```

And by site:

```{r}
matches %>%
  group_by(sink_site) %>%
  summarise(accuracy = sum(accurate) / n())
```

Overall, these are about what I would expect. The accuracy for public sites is
slightly better than random chance (which should produce ~25% accuracy), but
not by much.

### Effect of sampling delay

I'll explore the effect of sampling delay on matching accuracy. First I'll look
at the full set of matching attempts.

```{r}
matches %>%
  mutate(delay = sink_timepoint - source_timepoint) %>%
  group_by(delay) %>%
  summarise(accuracy = sum(accurate) / n()) %>%
  ungroup() %>%
  ggplot(aes(x = delay, y = accuracy)) +
    geom_line() +
    geom_smooth(method = "loess") +
    labs(title = "SourceTracker matching accuracy by sampling delay",
         subtitle = "For all query samples",
         caption = "Sampling delay is query time point minus reference time point",
         x = "Sampling delay",
         y = "Matching accuracy")
```

There's little clear effect of delay on matching accuracy.

I'll visualise the same results, but further broken down by query site.

```{r}
matches %>%
  mutate(delay = sink_timepoint - source_timepoint) %>%
  group_by(sink_site, delay) %>%
  summarise(accuracy = sum(accurate) / n()) %>%
  ungroup() %>%
  mutate_at("sink_site", str_to_sentence) %>%
  ggplot(aes(x = delay, y = accuracy)) +
    geom_line() +
    geom_smooth(method = "loess") +
    facet_wrap(~ sink_site, ncol = 1) +
    labs(title = "SourceTracker matching accuracy by sampling delay",
         subtitle = "Broken down by query site",
         caption = "Sampling delay is query time point minus reference time point",
         x = "Sampling delay",
         y = "Matching accuracy")
```

Again, if there is an effect here, it's very small.

### Diurnality effect

I'll now examine whether the time of day at which the query and reference
samples were taken affects matching accuracy.

```{r}
matches %>%
  left_join(metadata %>%
              select(source_timepoint = sink_timepoint, source_time = sink_time) %>% 
              distinct()) %>%
  group_by(sink_site, source_time, sink_time) %>% 
  summarise(accuracy = sum(accurate) / n()) %>%
  ungroup() %>%
  mutate_at(c("sink_site", "sink_time", "source_time"), str_to_sentence) %>%
  mutate(accuracy_lab = str_c(signif(accuracy, 2), "%")) %>%
  ggplot(aes(x = sink_site, y = accuracy, fill = sink_time, label = accuracy_lab)) +
    geom_col(position = "dodge") +
    geom_bar_text(position = "dodge") +
    facet_wrap(~ source_time, ncol = 1) +
    labs(title = "SourceTracker matching accuracy",
         subtitle = "Broken down by query site, query time and source time",
         x = "Query site",
         y = "Accuracy",
         fill = "Query time")
```

For the household sites at least, this is broadly consistent with my results from
the Watanabe method matching.
