---
title: Prepare inputs for SourceTracker-based microbiota matching
author: David Wilkins
date: Last updated `r lubridate::today()`
output:
  html_document:
    toc: true
    theme: readable
---

```{r}
library(tidyverse)
library(fs)
```

For SourceTracker-based microbiota matching, I want to generate predictions for
every surface sample with every skin sample as a potential source. Each run
will include all surface samples as sinks, and all skin samples from a given
time point as potential sources.

# Generating map files

As with the other matching methods, I'm only going to run potential matches
where there are a full set of four individuals in the reference pool. This
means time points that lack at least one skin sample from each individual will
not be run.

```{r}
if (! dir_exists("maps")) {
  dir_create("maps")
} else {
  dir_delete("maps")
  dir_create("maps")
}

metadata <- read_tsv("../metadata/metadata.tsv")

sinks <- metadata %>%
  filter(! type == "skin") %>%
  select("#SampleID" = sample, Env = type) %>%
  mutate(SourceSink = "sink")

runs <- metadata %>%
  filter(type == "skin") %>%
  group_by(timepoint) %>%
  mutate(pool_size = n_distinct(location)) %>%
  ungroup() %>%
  filter(pool_size == 4) %>%
  select(timepoint, "#SampleID" = sample, Env = location) %>%
  mutate(SourceSink = "source") %>%
  nest(map = c("#SampleID", Env, SourceSink)) %>%
  mutate(map = map(map, bind_rows, sinks)) %>%
  mutate(map_path = path("maps", timepoint, ext = "tsv"))

walk2(runs$map, runs$map_path, write_tsv)
```

# Generating OTU tables

Produce the OTU tables. Anything with an abundance smaller than 0.1% will be
discarded.

SourceTracker wants the OTU tables in a very specific format: the first line of
the file must be "#OTU TABLE"; the first column must be named "#OTU ID"; it
wants species in rows, samples in columns; and abundances must be given as
whole-number counts (I'll multiply the relative abundances by 1000 to achieve
this).

```{r}
if (! dir_exists("OTU_tables")) {
  dir_create("OTU_tables")
} else {
  dir_delete("OTU_tables")
  dir_create("OTU_tables")
}

abundances <- read_tsv("../abundances/species_abundances_decontaminated.tsv") %>%
  mutate(abundance = ceiling(1000 * abundance)) %>%
  filter(abundance > 0)

OTU_tables <- runs %>%
  mutate(samples = map(map, pull, "#SampleID")) %>%
  select(timepoint, samples) %>%
  mutate(OTU_table = list(abundances)) %>%
  mutate(OTU_table = map2(OTU_table, samples, ~ filter(.x, sample %in% .y))) %>%
  mutate(OTU_table = map(OTU_table, ~ .x %>% 
                         select(sample, species, abundance) %>%
                         spread(sample, abundance, fill = 0) %>%
                         rename("#OTU ID" = species)
                         )) %>%
  mutate(OTU_table_path = path("OTU_tables", timepoint, ext = "tsv"))

walk(OTU_tables$OTU_table_path, ~ write_tsv(tibble("#OTU TABLE" = character()), .x))
walk2(OTU_tables$OTU_table, OTU_tables$OTU_table_path, write_tsv, append =
      TRUE, col_names = TRUE)
```
