---
title: Microbial matching using the Bray-Curtis method
author: David Wilkins
date: Last updated `r lubridate::today()`
output:
  html_document:
    toc: true
    theme: readable
---

```{r global_options, include = FALSE}
library(knitr)
opts_chunk$set(fig.width = 12, fig.height = 8, echo = TRUE,
               warning = FALSE, message = FALSE)
```

```{r message = FALSE}
library(tidyverse)
library(printr)
library(vegan)
library(broom)
library(gglmannotate)
```

The aim of this analysis is to test the microbiota matching method using
Bray-Curtis distances (similar to the method introduced by Watanabe et al.
[^watanabe] using Canberra distances between samples). Ultimately I would like
to present and compare in the manuscript the accuracy of three matching methods
– SourceTracker matching, Bray-Curtis based matching, and Canberra based
(Watanabe) matching – and see how the accuracy varies with sampling delay.

This method requires Bray-Curtis distances between all samples, so I'll begin
by calculating these for all samples pairwise. I'll also save them to file for
potential later use.

# Computing Bray-Curtis distances between all samples

The vegan community matrix wants species as columns and samples as rows.

```{r}
comm <- read_tsv("../abundances/species_abundances_decontaminated.tsv") %>%
  select(sample, species, abundance) %>%
  spread(species, abundance, fill = 0) %>%
  as.data.frame() %>%
  column_to_rownames("sample") %>%
  as.matrix() 

distances <- vegdist(comm, method = "bray", upper = TRUE) 

bc <- distances %>%
  tidy() %>%
  rename(query_sample = item1, reference_sample = item2) %>%
  mutate_at(c("query_sample", "reference_sample"), as.character)

write_tsv(bc, "bc_distances.tsv")
```

# Performing matching with the Bray-Curtis method

Watanabe et al. used the mean of the Canberra distances between a query sample
and a pool of reference samples (i.e. the set of skin samples for an
individual) to perform matching. The reference individual with the lowest mean
distance was defined as the match.

For the purposes of this study, I will create a pool of reference individuals
for each time point and reference site (left or right palm), then compare each
query sample against that pool. The reference sample with the smallest mean
Bray-Curtis distance compared to the query sample will be considered the match.

I'll begin by setting up the reference pools. Pools with fewer than four
samples (i.e. for which there was not a sample for each of the four individuals
at that site for that time point) will not be used.

```{r}
metadata <- read_tsv("../metadata/metadata.tsv")

references <- metadata %>%
  filter(type == "skin") %>%
  select(-type) %>%
  add_count(site, timepoint, name = "pool_size") %>%
  filter(pool_size == 4) %>%
  select(-pool_size) %>%
  rename_all(~ str_c("reference_", .))
```

For each query sample (i.e. surface sample), I'll find the Bray-Curtis distance
between the query sample and each sample in each reference pool. For each
matching attempt (i.e. combination of query and reference pool), I'll then
assign the match as the reference sample with the smallest Bray-Curtis distance
from the query. The match is considered accurate if the match location is the
same as the query location.

```{r}
matches <- bc %>%
  left_join(metadata %>% rename_all(~ str_c("query_", .))) %>%
  filter(query_type %in% c("household surface", "public surface")) %>%
  right_join(references) %>%
  group_by(query_sample, reference_site, reference_timepoint) %>%
  mutate(is_winner = distance == min(distance)) %>%
  ungroup() %>%
  filter(is_winner) %>%
  select(-is_winner) %>%
  mutate(is_accurate = query_location == reference_location)
```

I'll write these results to file for later use.

```{r}
write_tsv(matches, "bc_matches.tsv")
```

# Accuracy of Bray-Curtis matching

I want to know a few things about the results of Bray-Curtis matching:

- What is the overall accuracy rate?
- What are the accuracy rates broken down by site?
- Is there any effect of sampling delay (the amount of time the query sample
  was taken after the reference sample) on matching accuracy?

## Accuracy rates

I'll start with the overall accuracy rate:

```{r}
matches %>%
  summarise(accuracy = sum(is_accurate) / n())
```

This is the lowest overall accuracy of any method I've tried so far.

Next I'll look at the accuracy rate broken down by query type:

```{r}
matches %>%
  group_by(query_type) %>%
  summarise(accuracy = sum(is_accurate) / n())
```

And by site:

```{r}
matches %>%
  group_by(query_site) %>%
  summarise(accuracy = sum(is_accurate) / n())
```

Again, this method is generally less accurate than the other two.

## Effect of sampling delay on matching accuracy

I'll visualise the matching accuracy for each site and for different sampling
delays.

```{r}
matches %>%
  mutate(delay = query_timepoint - reference_timepoint) %>%
  group_by(query_site, delay) %>%
  summarise(accuracy = sum(is_accurate) / n()) %>%
  ungroup() %>%
  mutate_at("query_site", str_to_sentence) %>%
  ggplot(aes(x = delay, y = accuracy)) +
    geom_line() +
    geom_smooth(method = "loess") +
    facet_wrap(~ query_site, ncol = 1) +
    labs(title = "Bray-Curtis matching accuracy by sampling delay",
         subtitle = "Broken down by query site",
         caption = "Sampling delay is query time point minus reference time point",
         x = "Sampling delay",
         y = "Matching accuracy")
```

As with the other two methods, there is little clear evidence of an effect
here.

[^watanabe]: Watanabe H, Nakamura I, Mizutani S, et al. Minor taxa in human skin microbiome contribute to the personal identification. Takemoto K, ed. PLoS ONE. 2018;13(7):e0199947. doi:10.1371/journal.pone.0199947.
