---
title: Modelling dispersal between sites with Dynamic Bayesian Networks
author: David Wilkins
date: Last updated `r lubridate::today()`
output:
  html_document:
    toc: true
    theme: readable
---

```{r global_options, include = FALSE}
library(knitr)
opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r message = FALSE}
library(tidyverse)
library(printr)
library(fs)
library(glue)
library(wilkoxmisc)
library(ggtext)
library(tidygraph)
library(ggraph)
library(broom)
```

The aim of this work is to model the dispersal of microorganisms between public
surface, indoor surface and skin sites using Dynamic Bayesian Networks. This
may give some insight into dispersal patterns. Broadly, I am following the
approach of Lax *et al.*[^lax], using BANJO to build the networks.

# Preparing data and generating DBNs

Following Lax *et al.*, I'm going to start with abundances aggregated at the
family level and log2-transformed although I might refine this as I go.

```{r}
walk(c("BANJO_input", "BANJO_output", "BANJO_settings"), function(d) {
  if (! dir_exists(d)) dir_create(d)
})

settings_template <- read_file("dynamic.settings.txt")

write_settings_file <- function(run_name, location, family, nvars, nobs, settings_file, ...) {
  settings <- glue(settings_template)
  write_file(settings, settings_file)
}

write_input_file <- function(observations, input_file, ...) {
  write_tsv(observations, input_file)
}

runs <- read_tsv("../abundances/species_abundances_decontaminated.tsv") %>%
  count(sample, site, type, location, timepoint, family, 
        wt = abundance, name = "abundance") %>%
  mutate(abundance = log2(abundance)) %>%
  mutate(abundance = if_else(is.infinite(abundance), 0, abundance)) %>%
  mutate_at(c("site", "location"), str_replace_all, " ", "-") %>%
  select(site, location, timepoint, family, abundance) %>%
  arrange(timepoint) %>%
  nest(observations = c(timepoint, site, abundance)) %>%
  mutate(observations = map(observations, pivot_wider, names_from = "site", 
                            values_from = "abundance", values_fill = list(abundance = 0))) %>%
  mutate(observations = map(observations, select, -timepoint)) %>%
  mutate(run_name = str_c(location, "_", family)) %>%
  mutate(settings_file = path("BANJO_settings", run_name, ext = "txt")) %>%
  mutate(input_file = path("BANJO_input", run_name, ext = "txt")) %>%
  mutate(output_file = path("BANJO_output", run_name, ext = "txt")) %>%
  mutate(nvars = map_dbl(observations, ncol)) %>%
  mutate(nobs = map_dbl(observations, nrow)) %>%
  pwalk(write_settings_file) %>%
  pwalk(write_input_file) %>%
  filter(! file_exists(output_file))

walk(
  runs$settings_file,
  ~ system(str_c("java -jar banjo.jar settingsFile=", .x))
)
```

# Load and parse output

```{r}
extract_graph <- function(output_file) {

  output <- read_file(output_file)

  # Extract the node labels
  node_sites <- str_match_all(output, "(\\d+) \\[label=\"(.+)\"\\]") %>%
    .[[1]] %>%
    as_tibble(.name_repair = "minimal") %>%
    set_names("original", "node", "site") %>%
    select(-original) %>%
    mutate(site = str_replace_all(site, "-", " ")) %>%
    distinct()

  # Extract the edges and translate to sites
  graph <- output %>%
    str_match_all("(\\d+)->(\\d+);") %>%
    .[[1]] %>%
    as_tibble(.name_repair = "minimal") %>%
    set_names("original", "parent", "child") %>%
    select(-original) %>%
    left_join(node_sites %>% rename(parent = node, parent_site = site), by = "parent") %>%
    left_join(node_sites %>% rename(child = node, child_site = site), by = "child") %>%
    select(parent_site, child_site) %>%
    distinct()
  return(graph)
}

graphs <- tibble(output_file = dir_ls("BANJO_output")) %>%
  mutate(location = str_extract(output_file, "residence-\\d")) %>%
  mutate(family = str_match(output_file, "residence-\\d_(.+)\\.txt")[,2]) %>%
  mutate_at(c("location", "family"), ~ str_replace_all(., "-", " ")) %>%
  mutate(graph = map(output_file, extract_graph))
```

# Summary statistics on DBN generation

Before unpacking the graphs themselves, I'm going to look a little closer at
the success rates for generating graphs. What proportion of runs generated a
graph?

```{r}
graphs %>%
  mutate(success = map_dbl(graph, nrow) > 0) %>%
  count(success) %>%
  mutate(percentage = 100 * n / sum(n))
```

A little fewer than half; not too bad. I'll look at how these success rates
distribute across locations and families.

```{r}
graphs %>%
  mutate(success = map_dbl(graph, nrow) > 0) %>%
  mutate_at("location", str_to_sentence) %>%
  count(location, success) %>%
  ggplot(aes(x = location, y = n, fill = success)) +
    geom_col(position = "stack") +
    labs(title = "DBN generation success by location",
         subtitle = "For runs at the family level",
         x = "Location",
         y = "Graphs",
         fill = "Success?")
```

Some variation, but promisingly every location succeeded in producing at least
some graphs.

```{r}
graphs %>%
  mutate(success = map_dbl(graph, nrow) > 0) %>%
  mutate_at("family", binom) %>%
  count(family, success) %>%
  mutate_at("family", fct_rev) %>%
  ggplot(aes(x = family, y = n, fill = success)) +
    geom_col(position = "stack") +
    labs(title = "DBN generation success by family",
         subtitle = "For runs at the family level",
         x = "Family",
         y = "Graphs",
         fill = "Success?") +
    theme(axis.text.y = element_markdown()) +
    coord_flip()
```

Not too bad. The majority of families produced at least some valid graphs.

# Summary statistics on DBNs

Next, I'll unpack the graphs to start looking at the graphs themselves.

```{r}
graphs <- graphs %>%
  unnest(graph) %>%
  select(-output_file)

count(graphs, parent_site, child_site) %>%
  arrange(desc(n))
```

I'll draw it as a directed graph.

```{r}
network <- graphs %>%
  count(parent_site, child_site) %>%
  as_tbl_graph()

ggraph(network) + 
    geom_edge_fan(aes(alpha = n, edge_width = n), arrow = grid::arrow()) + 
    geom_node_point() +
    geom_node_label(aes(label = name))
```

Grossly, this looks like surprisingly good support for dispersal from public
surfaces to skin. I'll get some edge statistics by type:

```{r}
site_type <- "../metadata/metadata.tsv" %>% 
  read_tsv() %>% 
  distinct(site, type)

graphs %>%
  left_join(site_type %>% rename(parent_site = site, parent_type = type)) %>%
  left_join(site_type %>% rename(child_site = site, child_type = type)) %>%
  count(parent_type, child_type) %>%
  arrange(desc(n))

graphs %>%
  left_join(site_type %>% rename(parent_site = site, parent_type = type)) %>%
  left_join(site_type %>% rename(child_site = site, child_type = type)) %>%
  count(parent_type) %>%
  arrange(desc(n))

graphs %>%
  left_join(site_type %>% rename(parent_site = site, parent_type = type)) %>%
  left_join(site_type %>% rename(child_site = site, child_type = type)) %>%
  count(child_type)
```

Overall, this nicely confirms and extends the findings of Lax et al. Skin is
more likely to be a parent than household surfaces, but public surfaces
outnumber them both. Public surfaces are also much less likely to be children
than either skin or household surfaces.

# Statistical significance

```{r}
graphs %>%
  left_join(site_type %>% rename(parent_site = site, parent_type = type)) %>%
  left_join(site_type %>% rename(child_site = site, child_type = type)) %>%
  count(location, parent_type, child_type) %>%
  ggplot(aes(x = parent_type, y = n)) +
    geom_boxplot() +
    facet_wrap(~ child_type)
```

```{r}
graphs %>%
  left_join(site_type %>% rename(parent_site = site, parent_type = type)) %>%
  left_join(site_type %>% rename(child_site = site, child_type = type)) %>%
  count(location, parent_type, child_type) %>%
  nest(data = c(parent_type, location, n)) %>%
  mutate(test = map(data, ~ kruskal.test(n ~ parent_type, data = .x))) %>%
  mutate(test = map(test, tidy)) %>%
  unnest(test)
```

```{r}
graphs %>%
  left_join(site_type %>% rename(parent_site = site, parent_type = type)) %>%
  left_join(site_type %>% rename(child_site = site, child_type = type)) %>%
  count(location, parent_type, child_type) %>%
  ggplot(aes(x = child_type, y = n)) +
    geom_boxplot() +
    facet_wrap(~ parent_type)
```

```{r}
graphs %>%
  left_join(site_type %>% rename(parent_site = site, parent_type = type)) %>%
  left_join(site_type %>% rename(child_site = site, child_type = type)) %>%
  count(location, parent_type, child_type) %>%
  nest(data = c(child_type, location, n)) %>%
  mutate(test = map(data, ~ kruskal.test(n ~ child_type, data = .x))) %>%
  mutate(test = map(test, tidy)) %>%
  unnest(test)
```

[^lax]: Lax S, Smith DP, Hampton-Marcell J, et al. Longitudinal analysis of
microbial interaction between humans and the indoor environment. Science.
2014;345(6200):1048-1052. doi:10.1126/science.1254529.
