---
title: Compare accuracy of microbial matching methods
author: David Wilkins
date: Last updated `r lubridate::today()`
output:
  html_document:
    toc: true
    theme: readable
---

```{r global_options, include = FALSE}
library(knitr)
opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r message = FALSE}
library(tidyverse)
library(printr)
library(broom)
library(gglmannotate)
```

The aim of this analysis is to collate the results of microbial matching using
the three methods (SourceTracker, Watanabe/Canberra, and Bray-Curtis). I want
to produce three things: a figure for the manuscript comparing the methods;
summary statistics of the accuracy rates for the three methods; and statistical
tests comparing the rates between the methods.

## Load and collate matching results

```{r}
st <- read_tsv("../st_matching/st_matches.tsv") %>%
  left_join(read_tsv("../metadata/metadata.tsv") %>%
                       select(timepoint, day, time) %>%
                       distinct() %>%
                       rename_all(~ str_c("source_", .))) %>%
  rename_all(~ str_replace(., "sink", "query")) %>%
  rename_all(~ str_replace(., "source", "reference")) %>%
  mutate(method = "SourceTracker") %>%
  mutate(reference_site = "pooled") %>%
  select(method, query_sample, query_site, query_type, query_location,
         query_day, query_time, query_timepoint, reference_site,
         reference_location, reference_day, reference_time, reference_timepoint,
         is_accurate = accurate)

watanabe <- read_tsv("../watanabe_matching/watanabe_matches.tsv") %>%
  mutate(method = "Canberra distance") %>%
  select(method, query_sample, query_site, query_type, query_location,
         query_day, query_time, query_timepoint, reference_site,
         reference_location, reference_day, reference_time, reference_timepoint,
         is_accurate)

bc <- read_tsv("../bc_matching/bc_matches.tsv") %>%
  mutate(method = "Bray-Curtis dissimilarity") %>%
  select(method, query_sample, query_site, query_type, query_location,
         query_day, query_time, query_timepoint, reference_site,
         reference_location, reference_day, reference_time, reference_timepoint,
         is_accurate)

matching <- bind_rows(st, watanabe, bc) %>%
  mutate(query_hours = (24 * (query_day - 1)) + if_else(query_time == "morning", 0, 12)) %>%
  mutate(reference_hours = (24 * (reference_day - 1)) + if_else(reference_time == "morning", 0, 12)) %>%
  mutate(delay = query_hours - reference_hours)
```

## Prepare matching accuracy figure

```{r}
matching %>%
  group_by(method, query_site, delay) %>%
  summarise(accuracy = sum(is_accurate) / n()) %>%
  ungroup() %>%
  mutate_at("accuracy", ~ . * 100) %>%
  mutate_at("query_site", str_to_sentence) %>%
  ggplot(aes(x = delay, y = accuracy, colour = method)) +
    geom_point() +
    geom_smooth(method = "loess") +
    facet_wrap(~ query_site, nrow = 1) +
    labs(x = "Sampling delay (hours)",
         y = "Accuracy (%)",
         colour = "Method") +
    theme_classic()

ggsave("matching_figure.png", width = 8, height = 3, dpi = 300)
```

I want to know: for how many combinations of query site and sampling delay is
the Canberra method the best-performing?

```{r}
matching %>%
  group_by(method, query_site, delay) %>%
  summarise(accuracy = sum(is_accurate) / n()) %>%
  ungroup() %>%
  group_by(query_site, delay) %>%
  mutate(is_winner = accuracy == max(accuracy)) %>%
  ungroup() %>%
  filter(is_winner) %>%
  select(-is_winner) %>%
  count(method)
```

## Accuracy rates by method and surface type

```{r}
matching %>%
  group_by(method, query_site, delay) %>%
  summarise(accuracy = sum(is_accurate) / n()) %>%
  ungroup() %>%
  left_join(read_tsv("../metadata/metadata.tsv") %>% 
              select(site, type) %>%
              distinct() %>%
              rename_all(~ str_c("query_", .))) %>%
  group_by(method, query_type) %>%
  summarise(mean_accuracy = mean(accuracy), sd_accuracy = sd(accuracy)) %>%
  ungroup()
```

## Effect of sampling delay on matching accuracy

First, I'm going to plot the relationship between absolute sampling delay and
matching accuracy. I'm only interested in Canberra matching going forward.

```{r}
matching %>%
  filter(method == "Canberra distance") %>%
  mutate_at("delay", abs) %>%
  group_by(query_site, delay) %>%
  summarise(accuracy = sum(is_accurate) / n()) %>%
  ungroup() %>%
  mutate_at("accuracy", ~ . * 100) %>%
  mutate_at("query_site", str_to_sentence) %>%
  ggplot(aes(x = delay, y = accuracy)) +
    geom_point() +
    geom_smooth(method = "lm") +
    facet_wrap(~ query_site, ncol = 1, scales = "free_y") +
    geom_lmannotate() +
    labs(title = "Relationship between absolute sampling delay and matching accuracy",
         x = "Sampling delay (hours)",
         y = "Matching accuracy (%)")
```

I'll use Spearman's rho to test for a significant relationship between absolute
sampling delay and matching accuracy across all sites.

```{r}
matching %>%
  filter(method == "Canberra distance") %>%
  select(query_site, delay, is_accurate) %>%
  mutate_at("delay", abs) %>%
  group_by(query_site, delay) %>%
  summarise(accuracy = sum(is_accurate) / n()) %>%
  ungroup() %>%
  arrange(delay) %>%
  nest(data = c(delay, accuracy)) %>%
  mutate(cor = map(data, ~ cor.test(.x$accuracy, .x$delay, method = "spearman"))) %>%
  mutate_at("cor", map, tidy) %>%
  select(-data) %>%
  unnest(cor) %>%
  mutate(significant = p.value < 0.05)
```

## Test for significant difference in accuracy by method for each query type

```{r}
matching %>%
  count(is_accurate, query_type, method) %>%
  spread(is_accurate, n) %>%
  nest(ct = c(method, `FALSE`, `TRUE`)) %>%
  mutate_at("ct", map, column_to_rownames, "method") %>%
  mutate_at("ct", map, as.matrix) %>%
  mutate(cs = map(ct, chisq.test)) %>%
  select(-ct) %>%
  mutate_at("cs", map, tidy) %>%
  unnest(cs) %>%
  mutate(is_significant = p.value < 0.05)
```
