---
title: Effect of diurnal variation on matching accuracy
author: David Wilkins
date: Last updated `r lubridate::today()`
---

```{r global_options, include = FALSE}
library(knitr)
opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r message = FALSE}
library(tidyverse)
library(printr)
library(broom)
library(glue)
library(ggfittext)
```

The purpose of this analysis is to test whether the diurnal variation in
species abundances on skin has any effect on matching accuracy. In other words,
for the purposes of microbiota matching, does it matter whether the reference
sample was taken in the morning or the evening?

## Matching accuracy

I'll begin by loading the matching results using the Watanabe method.

```{r}
matches <- read_tsv("../watanabe_matching/watanabe_matches.tsv")
```

I'll now break down the accuracy rates for morning and evening reference pools.
Since this figure will appear in the manuscript, I'll prepare it at publication
quality.

```{r}
matches %>%
  group_by(reference_time, query_type, query_time) %>%
  summarise(accuracy = 100 * (sum(is_accurate) / n())) %>%
  ungroup() %>%
  mutate(accuracy = signif(accuracy, 2)) %>%
  mutate_at(c("query_type", "reference_time", "query_time"), 
            str_to_sentence) %>%
  mutate(query_type = fct_relabel(query_type, ~ str_wrap(.x, width = 8))) %>%
  mutate_at(c("reference_time"), ~ glue("{.} reference")) %>%
  mutate_at(c("query_time"), ~ glue("{.} query")) %>%
  mutate_at(c("query_time", "reference_time"), ~ str_wrap(.x, width = 8)) %>%
  mutate_at(c("query_time", "reference_time"), fct_rev) %>%
  ggplot(aes(x = query_type, y = accuracy, label = glue("{accuracy}%"))) +
    geom_col(position = "dodge") +
    facet_grid(reference_time ~ query_time) +
    geom_bar_text(position = "dodge", min.size = 0) +
    labs(
      x = "Query type",
      y = "Accuracy (%)"
    ) +
    theme_classic()

ggsave("matching_diurnality.png", width = 3.5, height = 3, dpi = 300)
```

### Summary of results

- Among the household surface sites, evening queries outperform morning queries.
- Among the public surface sites, morning queries outperform evening queries.
- Among the household surface sites, morning references outperform evening
  references.
- Among the public surface sites, evening references outperform morning
  references.

### Statistical tests of these results

I'll perform chi-square tests to confirm the above results, and quantify the
effect sizes.

Null hypothesis: among household surface query sites, query time and matching
accuracy are independent.

```{r}
matches %>%
  filter(query_type == "household surface") %>%
  count(query_time, is_accurate) %>%
  spread(is_accurate, n) %>%
  as.data.frame() %>%
  column_to_rownames("query_time") %>%
  as.matrix() %>%
  chisq.test() %>%
  tidy()
```

Conclusion: the null hypothesis is rejected at p < 0.05.

I'll now quantify the effect:

```{r}
matches %>%
  filter(query_type == "household surface") %>%
  count(query_time, is_accurate) %>%
  spread(is_accurate, n) %>%
  mutate(accuracy = 100 * `TRUE` / (`TRUE` + `FALSE`))
```

**Statement: among household surface sites, evening queries have significantly higher
accuracy than morning queries.**

---

Null hypothesis: among public surface query sites, query time and matching
accuracy are independent.

```{r}
matches %>%
  filter(query_type == "public surface") %>%
  count(query_time, is_accurate) %>%
  spread(is_accurate, n) %>%
  as.data.frame() %>%
  column_to_rownames("query_time") %>%
  as.matrix() %>%
  chisq.test() %>%
  tidy()
```

Conclusion: the null hypothesis is rejected at p < 0.05.

I'll now quantify the effect:

```{r}
matches %>%
  filter(query_type == "public surface") %>%
  count(query_time, is_accurate) %>%
  spread(is_accurate, n) %>%
  mutate(accuracy = 100 * `TRUE` / (`TRUE` + `FALSE`))
```

**Statement: among public surface query sites, morning queries have significantly
higher accuracy than evening queries.**

---

Null hypothesis: among household surface query sites, reference time and matching
accuracy are independent.

```{r}
matches %>%
  filter(query_type == "household surface") %>%
  count(reference_time, is_accurate) %>%
  spread(is_accurate, n) %>%
  as.data.frame() %>%
  column_to_rownames("reference_time") %>%
  as.matrix() %>%
  chisq.test() %>%
  tidy()
```

Conclusion: the null hypothesis is rejected at p < 0.05.

I'll now quantify the effect:

```{r}
matches %>%
  filter(query_type == "household surface") %>%
  count(reference_time, is_accurate) %>%
  spread(is_accurate, n) %>%
  mutate(accuracy = 100 * `TRUE` / (`TRUE` + `FALSE`))
```

**Statement: among household surface query sites, morning references are significantly
more accurate than evening references.**

---

Null hypothesis: among public surface query sites, reference time and matching
accuracy are independent.

```{r}
matches %>%
  filter(query_type == "public surface") %>%
  count(reference_time, is_accurate) %>%
  spread(is_accurate, n) %>%
  as.data.frame() %>%
  column_to_rownames("reference_time") %>%
  as.matrix() %>%
  chisq.test() %>%
  tidy()
```

Conclusion: the null hypothesis is not rejected at p < 0.05.

**Statement: among public surface query sites, there is no significant difference in accuracy between morning and evening references.**

### Hypothesis arising from these results

In the morning, people have just finished a long period of interaction with
their household microbiome: arriving home from work, evening activities,
sleeping, and morning activities. Conversely, in the evening people have just
finished a long period of interaction with their local public microbiomes: out
and about on the subway, interacting with other people and with surfaces
bearing common (pooled) microbiomes. 

So, their morning microbiome more closely resembles their household microbiome,
while their evening microbiome more closely resembles their public
environmental microbiome.

Let's see how well this fits with the results above:

- Evening queries outperform morning queries for household surface sites
  - ✗ Why should an evening query, taken after a long period where the occupant
    has been absent, outperform a morning query, taken after a long period
    where the occupant has been present and presumably dispersing to household
    surfaces?
- Morning references outperform evening references for household surface sites
  - ✓ This fits with the idea that the public microbiome is 'erased' in favour
    of the household/personal one overnight

- Morning queries outperform evening queries for public surface sites
  - ✗ If the assumption is that public surfaces are relatively stable over
    time, why should there be a time-of-day effect for query samples? Also, the
    sampling delay effect tests show that it's precisely public surfaces that
    are *not* stable over time, while household surfaces are.
- There is no effect of reference time of day for public surface sites
  - ✓ This fits with the idea that the public microbiome disperses onto skin
    during interactions during the day time

Overall it looks like my model is missing something.
