---
title: Analyse results of SourceTracker predictions for surface sources
author: David Wilkins
date: Last updated `r lubridate::today()`
output:
  html_document:
    toc: true
    theme: readable
---

```{r global_options, include = FALSE}
library(knitr)
opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r message = FALSE}
library(tidyverse)
library(printr)
library(fs)
```

The aim of this analysis is to examine the results of the SourceTracker
predictions for surfaces. This is part of my broader goal of confirming human
skin as the major source for both household and public surface microbiota, as
well as checking that the negative controls were not a major source.

# Load and tidy SourceTracker source predictions

```{r}
metadata <- read_tsv("../metadata/metadata.tsv") %>%
  rename_all(~ str_c("sink_", .))

source_predictions <- tibble(sp_dir = dir_ls("source_predictions")) %>%
  mutate(run = path_file(sp_dir)) %>%
  mutate_at("run", str_replace, "-", " ") %>%
  separate(run, c("source_location", "run_timepoint"), "_") %>%
  mutate(sp_path = path(sp_dir, "sink_predictions", ext = "txt")) %>%
  mutate(sp = map(sp_path, read_tsv)) %>%
  select(source_location, sp) %>%
  unnest(sp) %>%
  rename(sink_sample = SampleID) %>%
  gather("source", "proportion", -source_location, -sink_sample) %>%
  left_join(metadata)
```

I'll write the source predictions to file for use later.

```{r}
write_tsv(source_predictions, "surface_source_predictions.tsv")
```

# Visualise sources for each surface site

I'll plot out the source proportions across each site for all surface samples.

```{r}
source_predictions %>%
  mutate_at(c("source", "sink_site"), str_to_sentence) %>%
  mutate_at("source", fct_relevel, "Control", "Unknown", "Skin") %>%
  ggplot(aes(x = source, y = proportion)) +
    geom_boxplot() +
    facet_wrap(~ sink_site, nrow = 1) +
    labs(title = "SourceTracker predicted sources for all surface sites",
         x = "Source",
         y = "Predicted contribution") +
    coord_flip()
```

This confirms that human skin is the major identifiable source for all
household and public surface sites.

# Quantification of skin vs non-skin contribution to surface types

I'll calculate the mean and SD of the predicted source proportions for each
type so I can report them in the manuscript.

```{r}
source_predictions %>%
  select(sink_sample, sink_type, source, proportion) %>%
  complete(nesting(sink_sample, sink_type), source, fill = list(proportion = 0)) %>%
  group_by(sink_type, source) %>%
  summarise(mean_proportion = mean(proportion), sd_proportion = sd(proportion)) %>%
  ungroup()
```
