---
title: Prepare SourceTracker inputs for runs to determine sources of surface microbiota
author: David Wilkins
date: Last updated `r lubridate::today()`
output:
  html_document:
    toc: true
    theme: readable
---

```{r}
library(tidyverse)
library(purrr)
library(fs)
```

This SourceTracker run will predict sources for household and public surface
samples. The aim of this is two-fold: to confirm skin as the major source for
surface microbiota, and to exclude negative control samples as a likely source.

For each sink sample, potential sources will be within-location occupant skin,
the negative control samples, and unknown. To minimise the total number of
runs, only sources from the same time point as the sink sample will be
considered.

# Prepare map files

I'll prepare a map file for each run. Each run represents a single time
location and time point. The sink samples for the run are all the surface
samples for that location and time point, while the source samples are a
combination of the skin samples from that location and time point and the
negative control pool.

Each map file wants three columns: #SampleID, Env, SourceSink.

```{r}
if (! dir_exists("maps")) {
  dir_create("maps")
} else {
  dir_delete("maps")
  dir_create("maps")
}

metadata <- read_tsv("../metadata/metadata.tsv")
control_metadata <- read_tsv("../metadata/negative_control_metadata.tsv")

skin_sources <- metadata %>%
  filter(type == "skin") %>%
  select(sample, type, location, timepoint) %>%
  rename_all(~ str_c("source_", .x))

control_sources <- control_metadata %>%
  select(source_sample = sample) %>%
  mutate(source_type = "control")

sinks <- metadata %>%
  filter(! type == "skin") %>%
  select(sample, type, location, timepoint) %>%
  rename_all(~ str_c("sink_", .x))

runs <- metadata %>%
  distinct(location, timepoint) %>%
  mutate(sources = list(skin_sources)) %>%
  mutate(sources = map2(sources, location, ~ filter(.x, source_location == .y))) %>%
  mutate(sources = map2(sources, timepoint, ~ filter(.x, source_timepoint == .y))) %>%
  mutate(sources = map(sources, ~ select(.x, source_sample, source_type))) %>%
  mutate(skin_sources = map_dbl(sources, nrow)) %>%
  filter(skin_sources > 0) %>%
  select(-skin_sources) %>%
  mutate(sources = map(sources, ~ bind_rows(.x, control_sources))) %>%
  mutate(sinks = list(sinks)) %>%
  mutate(sinks = map2(sinks, location, ~ filter(.x, sink_location == .y))) %>%
  mutate(sinks = map2(sinks, timepoint, ~ filter(.x, sink_timepoint == .y))) %>%
  mutate(sink_count = map_dbl(sinks, nrow)) %>%
  filter(sink_count > 0) %>%
  select(-sink_count) %>%
  mutate(sources = map(sources, ~ rename(.x, "#SampleID" = source_sample, Env = source_type))) %>%
  mutate(sources = map(sources, ~ mutate(.x, SourceSink = "source"))) %>%
  mutate(sinks = map(sinks, ~ select(.x, "#SampleID" = sink_sample, Env = sink_type))) %>%
  mutate(sinks = map(sinks, ~ mutate(.x, SourceSink = "sink"))) %>%
  mutate(map = map2(sources, sinks, bind_rows)) %>%
  select(-sources, -sinks) %>%
  mutate(run = str_c(location, timepoint, sep = "_")) %>%
  mutate_at("run", str_replace, " ", "-") %>%
  mutate(map_path = path("maps", run, ext = "tsv"))

walk2(runs$map, runs$map_path, write_tsv)
```

# OTU tables

Produce the OTU tables. Anything with an abundance smaller than 0.1% will be
discarded.

SourceTracker wants the OTH tables in a very specific format: the first line of
the file must be "#OTU TABLE"; the first column must be named "#OTU ID"; it
wants species in rows, samples in columns; and abundances must be given as
whole-number counts (I'll multiply the relative abundances by 1000 to achieve 
this).

```{r}
if (! dir_exists("OTU_tables")) {
  dir_create("OTU_tables")
} else {
  dir_delete("OTU_tables")
  dir_create("OTU_tables")
}

abundances <- read_tsv("../abundances/species_abundances_decontaminated.tsv") %>%
  mutate(abundance = floor(1000 * abundance)) %>%
  filter(abundance > 0) %>%
  select(sample, species, abundance)

negative_abundances <- read_tsv("../abundances/control_species_abundances.tsv") %>%
  mutate(abundance = floor(1000 * abundance)) %>%
  filter(abundance > 0)

abundances <- bind_rows(abundances, negative_abundances)

OTU_tables <- runs %>%
  mutate(samples = map(map, pull, "#SampleID")) %>%
  select(run, samples) %>%
  mutate(OTU_table = list(abundances)) %>%
  mutate(OTU_table = map2(OTU_table, samples, ~ filter(.x, sample %in% .y))) %>%
  mutate(OTU_table = map(OTU_table, ~ .x %>% 
                         select(sample, species, abundance) %>%
                         spread(sample, abundance, fill = 0) %>%
                         rename("#OTU ID" = species)
                         )) %>%
  mutate(OTU_table_path = path("OTU_tables", run, ext = "tsv"))


walk(OTU_tables$OTU_table_path, ~ write_tsv(tibble("#OTU TABLE" = character()), .x))
walk2(OTU_tables$OTU_table, OTU_tables$OTU_table_path, write_tsv, append =
      TRUE, col_names = TRUE)
```
